---
title: "Aurora Lending Group: Portfolio Risk Review (R)"
subtitle: "Consumer lending book, originations 2019-2025"
author: "Risk Analytics"
date: today
format:
  html:
    theme: cosmo
    brand: _brand.yml
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 9
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
engine: knitr
---

::: {.callout-important}
## Synthetic data

This project contains synthetic data and analysis created for demonstration
purposes only. Aurora Lending Group is a fictional company. No real borrower
data, portfolio data, or performance data went into this report.
:::

::: {.callout-note}
## AI-generated content

AI wrote the commentary and the business scenarios in this report, as an
illustration. No subject-matter expert reviewed them.
:::

This is the R parallel of `reports/aurora_lending/portfolio_risk_review.qmd`,
built with `tidyverse`, `gt`, and `tidymodels`' `yardstick` instead of
`pandas`, `great_tables`, and scikit-learn's metrics.

```{r}
#| label: setup
library(DBI)
library(dplyr)
library(ggplot2)
library(patchwork)
library(gt)
library(paws)
library(tidymodels)

# This report goes to Posit Connect as a self-contained document. It does not
# source config.R from the repository, because the bundle does not hold that
# file. The settings it needs are below. Each one reads an environment
# variable first, so Connect can point the report at another account.
# tests/test_report_config.py checks the Python report against config.py;
# keep this file in agreement with both by hand.
REGION <- Sys.getenv("POSIT_DEMO_REGION", "us-east-2")
WORKGROUP <- Sys.getenv("POSIT_DEMO_WORKGROUP", "primary")
DATABASE <- Sys.getenv("POSIT_DEMO_DATABASE", "aurora_lending")
ENDPOINT <- Sys.getenv("POSIT_DEMO_ENDPOINT", "aurora-lending-risk")
BUCKET_PREFIX <- "sagemaker-posit-conf-2026-demo"

# The driver name and the authentication type both differ between drivers.
# The defaults are what the Positron SageMaker image uses: the Posit
# Professional Drivers register as "Athena", and they authenticate with the
# execution role.
ATHENA_DRIVER <- Sys.getenv("POSIT_DEMO_ATHENA_DRIVER", "Athena")
ATHENA_AUTH <- Sys.getenv("POSIT_DEMO_ATHENA_AUTH", "Default")

sts <- paws::sts(config = list(region = REGION))

# This report asks STS for the account number, and does not store it. This
# public repository therefore holds no account number, and a fork needs no edit.
BUCKET <- Sys.getenv("POSIT_DEMO_BUCKET")
if (!nzchar(BUCKET)) {
  BUCKET <- sprintf("%s-%s-%s", BUCKET_PREFIX, sts$get_caller_identity()$Account, REGION)
}
ATHENA_STAGING <- sprintf("s3://%s/athena-query-results/", BUCKET)

CONNECTION_STRING <- paste0(
  "Driver=", ATHENA_DRIVER, ";",
  "AwsRegion=", REGION, ";",
  "S3OutputLocation=", ATHENA_STAGING, ";",
  "AuthenticationType=", ATHENA_AUTH, ";",
  "Workgroup=", WORKGROUP, ";",
  "Schema=", DATABASE
)

POSIT_BLUE <- "#447099"
POSIT_ORANGE <- "#ee6331"
POSIT_BURGUNDY <- "#9a4665"

theme_set(theme_minimal(base_size = 9))

#' Read Athena over ODBC, with the driver the image ships.
athena <- function(sql) {
  con <- DBI::dbConnect(odbc::odbc(), .connection_string = CONNECTION_STRING)
  on.exit(DBI::dbDisconnect(con), add = TRUE)
  DBI::dbGetQuery(con, sql) |> as_tibble()
}
```

```{r}
#| label: load
loans <- athena("
    SELECT f.loan_id,
           f.origination_date,
           f.fico_at_origination,
           f.dti,
           f.apr,
           f.term_months,
           CAST(f.principal_amount AS double)    AS principal_amount,
           CAST(f.outstanding_balance AS double) AS outstanding_balance,
           f.employment_length_years,
           f.purpose,
           f.charged_off,
           b.state,
           b.income_band
    FROM fct_loan_performance f
    JOIN dim_borrower b ON b.borrower_id = f.borrower_id
")

loans <- loans |>
  mutate(
    vintage = as.integer(format(as.Date(origination_date), "%Y")),
    fico_band = cut(
      fico_at_origination,
      breaks = c(579, 620, 660, 700, 740, 780, 850),
      labels = c("580-619", "620-659", "660-699", "700-739", "740-779", "780+")
    )
  )

originated <- sum(loans$principal_amount)
outstanding <- sum(loans$outstanding_balance)
chargeoff_rate <- mean(loans$charged_off)
```

## Portfolio at a glance

The book holds **`{r} format(nrow(loans), big.mark = ",")`** loans
originated between `{r} min(loans$vintage)` and `{r} max(loans$vintage)`,
totaling **`{r} sprintf("$%.0fM", originated / 1e6)`** advanced and
**`{r} sprintf("$%.0fM", outstanding / 1e6)`** still outstanding. The
realized charge-off rate is **`{r} sprintf("%.1f%%", chargeoff_rate * 100)`**.

This report reads the data from Amazon Athena each time it renders. It
copies nothing and it caches nothing.

```{r}
#| label: tbl-purpose
by_purpose <- loans |>
  summarise(
    loans = n(),
    originated = sum(principal_amount),
    avg_apr = mean(apr),
    avg_fico = mean(fico_at_origination),
    chargeoff = mean(charged_off),
    .by = purpose
  ) |>
  arrange(desc(chargeoff)) |>
  mutate(purpose = tools::toTitleCase(gsub("_", " ", purpose)))

by_purpose |>
  gt() |>
  tab_header(title = "Performance by loan purpose",
             subtitle = "Ordered by realized charge-off rate") |>
  cols_label(purpose = "Purpose", loans = "Loans", originated = "Originated",
             avg_apr = "Avg APR", avg_fico = "Avg FICO", chargeoff = "Charge-off rate") |>
  fmt_number(columns = loans, decimals = 0, use_seps = TRUE) |>
  fmt_currency(columns = originated, currency = "USD", decimals = 0) |>
  fmt_percent(columns = c(avg_apr, chargeoff), decimals = 1) |>
  fmt_number(columns = avg_fico, decimals = 0) |>
  data_color(columns = chargeoff, palette = c("#ffffff", POSIT_ORANGE)) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_source_note("Synthetic data, generated for demonstration purposes.")
```

Small business lending has the highest loss rate. Auto refinance has the
lowest. The gap between them is large, so purpose is useful on its own as a
segment for a pricing review.

## Where the risk sits

```{r}
#| label: fig-risk
#| fig-cap: "Charge-off concentration by credit quality and origination vintage."
band <- loans |> summarise(rate = mean(charged_off) * 100, .by = fico_band)

p1 <- ggplot(band, aes(fico_band, rate)) +
  geom_col(fill = POSIT_BLUE, width = 0.72) +
  labs(title = "By FICO band at origination", x = NULL, y = "charge-off rate (%)")

vintage <- loans |> summarise(rate = mean(charged_off) * 100, .by = c(vintage, fico_band))

p2 <- ggplot(vintage, aes(vintage, rate, color = fico_band)) +
  geom_line(linewidth = 0.85) +
  geom_point() +
  scale_color_viridis_d(name = "FICO band") +
  labs(title = "By origination vintage", x = NULL, y = NULL)

p1 + p2
```

Credit quality is the strongest driver. The weakest band loses about
`{r} round(band$rate[1] / band$rate[nrow(band)])` times as much as the
strongest band. The newest vintages have less time on book, so their rates
are lower than the loss they will reach.

## Model-scored risk

```{r}
#| label: score
#| output: false
features <- c("fico_at_origination", "dti", "apr", "term_months",
              "principal_amount", "employment_length_years", "purpose")

set.seed(20260821)
sample_loans <- loans |> slice_sample(n = min(2000, nrow(loans)))
endpoint_available <- TRUE
endpoint_error <- ""

# The endpoint exists only around the time of a session. If it is gone, this
# report must still render on Connect. The error is therefore stored and
# shown on the page, and the render continues.
result <- tryCatch(
  {
    runtime <- paws::sagemakerruntime(config = list(region = REGION))
    body <- jsonlite::toJSON(list(instances = sample_loans |> select(all_of(features))),
                              dataframe = "rows")
    response <- runtime$invoke_endpoint(
      EndpointName = ENDPOINT, ContentType = "application/json", Body = charToRaw(body)
    )
    predictions <- jsonlite::fromJSON(rawToChar(response$Body))$predictions
    sample_loans <- sample_loans |> mutate(risk_score = predictions)
    NULL
  },
  error = function(err) err
)

if (!is.null(result)) {
  endpoint_available <- FALSE
  # AWS error text repeats the account number. This report gets published,
  # so the number is masked before it reaches the page.
  endpoint_error <- gsub("\\b\\d{12}\\b", "<account>",
                         sprintf("%s: %s", class(result)[1], conditionMessage(result)))
}
```

```{r}
#| label: score-note
#| output: asis
if (!endpoint_available) {
  cat(":::: {.callout-warning}\n\n")
  cat("## Model scores unavailable\n\n")
  cat(sprintf(
    "The SageMaker endpoint `%s` did not respond, so this section is omitted. Everything above is queried directly from Athena and is unaffected.\n\n",
    ENDPOINT
  ))
  cat("Deploy it with `uv run python ml/train_and_deploy.py --domain finance`, then re-render.\n\n")
  cat(sprintf("> `%s`\n\n", endpoint_error))
  cat("::::\n")
}
```

```{r}
#| label: tbl-lift
if (endpoint_available) {
  sample_loans <- sample_loans |> mutate(decile = ntile(risk_score, 10))
  lift <- sample_loans |>
    summarise(
      loans = n(), predicted = mean(risk_score), actual = mean(charged_off),
      exposure = sum(outstanding_balance), .by = decile
    ) |>
    arrange(desc(decile))

  lift |>
    gt() |>
    tab_header(title = "Predicted versus realized charge-off by risk decile",
               subtitle = sprintf("%s loans scored on endpoint %s",
                                   format(nrow(sample_loans), big.mark = ","), ENDPOINT)) |>
    cols_label(decile = "Decile", loans = "Loans", predicted = "Predicted",
               actual = "Realized", exposure = "Outstanding") |>
    fmt_number(columns = loans, decimals = 0, use_seps = TRUE) |>
    fmt_percent(columns = c(predicted, actual), decimals = 1) |>
    fmt_currency(columns = exposure, currency = "USD", decimals = 0) |>
    data_color(columns = actual, palette = c("#ffffff", POSIT_BURGUNDY)) |>
    tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
    tab_source_note("Decile 10 is the highest predicted risk. Computed with tidymodels' yardstick alongside this table. Synthetic data, generated for demonstration purposes.")
}
```

```{r}
#| label: summary
#| output: asis
if (endpoint_available) {
  auc <- sample_loans |>
    mutate(charged_off_fct = factor(charged_off, levels = c(1, 0))) |>
    yardstick::roc_auc(truth = charged_off_fct, risk_score, event_level = "first")

  base <- mean(sample_loans$charged_off) * 100
  top <- lift$actual[lift$decile == 10] * 100
  captured <- sum(sample_loans$charged_off[sample_loans$decile >= 9]) /
    max(sum(sample_loans$charged_off), 1) * 100
  at_risk <- sum(sample_loans$outstanding_balance[sample_loans$decile >= 9])

  cat(sprintf(
    "The model separates the book (yardstick AUC **%.3f**). The base rate for the portfolio is **%.1f%%**. The riskiest decile reaches **%.1f%%**, which is **%.1fx** the base rate.\n\n",
    auc$.estimate, base, top, top / base
  ))
  cat(sprintf(
    "The top two deciles hold **%.0f%%** of all charge-offs and **$%.1fM** of outstanding balance. Collections work and credit line reviews return the most in these two deciles.\n\n",
    captured, at_risk / 1e6
  ))
  cat("The predicted rate and the actual rate are close in every decile. You can therefore read the scores as rates, and not only as a rank order.")
}
```

## Method

A script makes the data. The data goes to Amazon S3 as parquet, and the
tables are registered in the AWS Glue Data Catalog.

This report reads those tables through Amazon Athena when it renders. It
then scores a sample of loans against a logistic risk model. Amazon
SageMaker hosts that model on a real-time endpoint. `tidymodels`' `yardstick`
computes the AUC from those scores; the model itself trains in Python
(`ml/train_and_deploy.py`) and is not refit here.

Every step runs from Positron on Amazon SageMaker, and the report is
published to Posit Connect.

::: {.callout-tip collapse="true"}
## Build this report again

```bash
uv run python data/generate.py     --domain finance
uv run python load/load_athena.py  --domain finance
uv run python ml/train_and_deploy.py --domain finance
quarto render reports/aurora_lending/portfolio_risk_review_r.qmd
uv run python ml/teardown.py --all
```
:::
