---
title: "Helix Therapeutics: Study HLX-301 Enrollment and Retention Review (R)"
subtitle: "Randomized, placebo-controlled Phase III"
author: "Clinical Data Science"
date: today
format:
  html:
    theme: cosmo
    brand: _brand.yml
    toc: true
    toc-depth: 3
    embed-resources: true
    fig-width: 9
    fig-height: 4.5
execute:
  echo: false
  warning: false
  message: false
engine: knitr
---

::: {.callout-important}
## Synthetic data

This project contains synthetic data and analysis created for demonstration
purposes only. Helix Therapeutics and study HLX-301 are fictional. No real
subject data, site data, or trial data went into this report.
:::

::: {.callout-note}
## AI-generated content

AI wrote the commentary and the clinical scenarios in this report, as an
illustration. No subject-matter expert reviewed them.
:::

This is the R parallel of
`reports/helix_trials/enrollment_safety_review.qmd`, built with `tidyverse`,
`gt`, and `tidymodels` instead of `pandas`, `great_tables`, and
scikit-learn's metrics.

```{r}
#| label: setup
library(DBI)
library(dplyr)
library(ggplot2)
library(patchwork)
library(gt)
library(paws)
library(tidymodels)

# This report goes to Posit Connect as a self-contained document. It does not
# source config.R from the repository, because the bundle does not hold that
# file. The settings it needs are below. Each one reads an environment
# variable first, so Connect can point the report at another account.
# tests/test_report_config.py checks the Python report against config.py;
# keep this file in agreement with both by hand.
REGION <- Sys.getenv("POSIT_DEMO_REGION", "us-east-2")
WORKGROUP <- Sys.getenv("POSIT_DEMO_WORKGROUP", "primary")
DATABASE <- Sys.getenv("POSIT_DEMO_DATABASE", "helix_trials")
ENDPOINT <- Sys.getenv("POSIT_DEMO_ENDPOINT", "helix-dropout-risk")
BUCKET_PREFIX <- "sagemaker-posit-conf-2026-demo"

ATHENA_DRIVER <- Sys.getenv("POSIT_DEMO_ATHENA_DRIVER", "Athena")
ATHENA_AUTH <- Sys.getenv("POSIT_DEMO_ATHENA_AUTH", "Default")

sts <- paws::sts(config = list(region = REGION))

BUCKET <- Sys.getenv("POSIT_DEMO_BUCKET")
if (!nzchar(BUCKET)) {
  BUCKET <- sprintf("%s-%s-%s", BUCKET_PREFIX, sts$get_caller_identity()$Account, REGION)
}
ATHENA_STAGING <- sprintf("s3://%s/athena-query-results/", BUCKET)

CONNECTION_STRING <- paste0(
  "Driver=", ATHENA_DRIVER, ";",
  "AwsRegion=", REGION, ";",
  "S3OutputLocation=", ATHENA_STAGING, ";",
  "AuthenticationType=", ATHENA_AUTH, ";",
  "Workgroup=", WORKGROUP, ";",
  "Schema=", DATABASE
)

POSIT_BLUE <- "#447099"
POSIT_ORANGE <- "#ee6331"
POSIT_BURGUNDY <- "#9a4665"

theme_set(theme_minimal(base_size = 9))

#' Read Athena over ODBC, with the driver the image ships.
athena <- function(sql) {
  con <- DBI::dbConnect(odbc::odbc(), .connection_string = CONNECTION_STRING)
  on.exit(DBI::dbDisconnect(con), add = TRUE)
  DBI::dbGetQuery(con, sql) |> as_tibble()
}
```

```{r}
#| label: load
# Group the visit history over a fixed early window first, Screening to Week
# 12, and then join to the subject table. If you join on a later visit
# first, the query removes every subject who already withdrew. That is
# survivorship bias, and it reports a dropout rate far below the true one.
subjects <- athena("
    WITH early_window AS (
        SELECT subject_id,
               count(*)                                                  AS early_visits,
               CAST(sum(CASE WHEN NOT visit_completed THEN 1 ELSE 0 END) AS double)
                   / count(*)                                            AS missed_visit_rate
        FROM fct_visit_observations
        WHERE visit_id <= 5
        GROUP BY subject_id
    )
    SELECT s.subject_id, s.arm, s.age, s.sex, s.ecog_status, s.bmi,
           s.prior_therapy_lines, s.baseline_das, s.early_ae_count,
           w.missed_visit_rate, s.discontinued,
           t.site_name, t.country, t.region
    FROM dim_subject s
    JOIN early_window w ON w.subject_id = s.subject_id
    JOIN dim_site t     ON t.site_id    = s.site_id
")

das <- athena("
    SELECT v.visit_name, v.scheduled_day, s.arm,
           avg(o.das_score) AS mean_das,
           count(*)         AS observations
    FROM fct_visit_observations o
    JOIN dim_subject s ON s.subject_id = o.subject_id
    JOIN dim_visit   v ON v.visit_id   = o.visit_id
    WHERE v.scheduled_day >= 0
    GROUP BY v.visit_name, v.scheduled_day, s.arm
    ORDER BY v.scheduled_day
")

n_subjects <- nrow(subjects)
n_sites <- n_distinct(subjects$site_name)
discont_rate <- mean(subjects$discontinued)
```

## Study status

**`{r} format(n_subjects, big.mark = ",")`** subjects are randomized across
**`{r} n_sites`** active sites in
**`{r} n_distinct(subjects$region)`** regions. Overall early discontinuation
stands at **`{r} sprintf("%.1f%%", discont_rate * 100)`**.

This report reads the data from Amazon Athena each time it renders.

```{r}
#| label: tbl-arms
arms <- subjects |>
  summarise(
    subjects = n(), mean_age = mean(age), baseline_das = mean(baseline_das),
    early_ae = mean(early_ae_count), discontinued = mean(discontinued),
    .by = arm
  )

arms |>
  gt() |>
  tab_header(title = "Balance and retention by treatment arm",
             subtitle = "Discontinuation computed over all randomized subjects") |>
  cols_label(arm = "Arm", subjects = "Subjects", mean_age = "Mean age",
             baseline_das = "Baseline DAS", early_ae = "Mean AEs to Wk 12",
             discontinued = "Discontinued") |>
  fmt_number(columns = subjects, decimals = 0, use_seps = TRUE) |>
  fmt_number(columns = c(mean_age, baseline_das, early_ae), decimals = 2) |>
  fmt_percent(columns = discontinued, decimals = 1) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
  tab_source_note("Synthetic data, generated for demonstration purposes.")
```

The two arms are balanced on age and on baseline disease activity. By Week
12 the treatment arm records more adverse events, and it also loses more
subjects. This is the expected trade-off for an active drug.

## Efficacy and safety

```{r}
#| label: fig-efficacy
#| fig-cap: "Disease activity trajectory by arm, and discontinuation against early adverse-event burden."
p1 <- ggplot(das, aes(scheduled_day, mean_das, color = arm)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  scale_color_manual(values = c(Treatment = POSIT_BLUE, Placebo = POSIT_ORANGE)) +
  labs(title = "Mean disease activity score (lower is better)",
       x = "study day", y = "DAS")

by_ae <- subjects |> summarise(rate = mean(discontinued) * 100, .by = early_ae_count)

p2 <- ggplot(by_ae, aes(factor(early_ae_count), rate)) +
  geom_col(fill = POSIT_BURGUNDY, width = 0.72) +
  labs(title = "Discontinuation by AE burden to Week 12",
       x = "adverse events recorded", y = "discontinued (%)")

p1 + p2
```

```{r}
#| label: efficacy-text
#| output: asis
wk24 <- das |> filter(scheduled_day == 168) |> select(arm, mean_das)
if (all(c("Treatment", "Placebo") %in% wk24$arm)) {
  treat <- wk24$mean_das[wk24$arm == "Treatment"]
  plac <- wk24$mean_das[wk24$arm == "Placebo"]
  sep <- plac - treat
  cat(sprintf(
    "At Week 24 the treatment arm is at **%.2f** and placebo is at **%.2f**. The gap is **%.2f DAS points**, and it holds to the end of the study.\n\n",
    treat, plac, sep
  ))
  cat("Discontinuation rises quickly as early adverse events increase. This is where a study team can act. The adverse event count at Week 12 is visible long before a subject withdraws.")
}
```

## Dropout risk

```{r}
#| label: score
#| output: false
features <- c("age", "ecog_status", "bmi", "prior_therapy_lines",
              "baseline_das", "early_ae_count", "missed_visit_rate", "arm")

endpoint_available <- TRUE
endpoint_error <- ""

result <- tryCatch(
  {
    runtime <- paws::sagemakerruntime(config = list(region = REGION))
    body <- jsonlite::toJSON(list(instances = subjects |> select(all_of(features))),
                              dataframe = "rows")
    response <- runtime$invoke_endpoint(
      EndpointName = ENDPOINT, ContentType = "application/json", Body = charToRaw(body)
    )
    predictions <- jsonlite::fromJSON(rawToChar(response$Body))$predictions
    subjects <- subjects |> mutate(dropout_risk = predictions)
    NULL
  },
  error = function(err) err
)

if (!is.null(result)) {
  endpoint_available <- FALSE
  endpoint_error <- gsub("\\b\\d{12}\\b", "<account>",
                         sprintf("%s: %s", class(result)[1], conditionMessage(result)))
}
```

```{r}
#| label: score-note
#| output: asis
if (!endpoint_available) {
  cat(":::: {.callout-warning}\n\n")
  cat("## Dropout scores unavailable\n\n")
  cat(sprintf(
    "The SageMaker endpoint `%s` did not respond, so this section is omitted. Everything above is queried directly from Athena and is unaffected.\n\n",
    ENDPOINT
  ))
  cat("Deploy it with `uv run python ml/train_and_deploy.py --domain lifesci`, then re-render.\n\n")
  cat(sprintf("> `%s`\n\n", endpoint_error))
  cat("::::\n")
}
```

```{r}
#| label: tbl-sites
if (endpoint_available) {
  watchlist <- subjects |>
    summarise(
      enrolled = n(), predicted = mean(dropout_risk), actual = mean(discontinued),
      .by = c(region, site_name)
    ) |>
    filter(enrolled >= 20) |>
    arrange(desc(predicted)) |>
    head(10)

  watchlist |>
    gt() |>
    tab_header(title = "Sites with the highest predicted dropout risk",
               subtitle = "Sites with at least 20 subjects enrolled") |>
    cols_label(region = "Region", site_name = "Site", enrolled = "Enrolled",
               predicted = "Predicted risk", actual = "Discontinued to date") |>
    fmt_number(columns = enrolled, decimals = 0) |>
    fmt_percent(columns = c(predicted, actual), decimals = 1) |>
    data_color(columns = predicted, palette = c("#ffffff", POSIT_BURGUNDY)) |>
    tab_style(style = cell_text(weight = "bold"), locations = cells_column_labels()) |>
    tab_source_note(sprintf("Scored on endpoint %s. Synthetic data, generated for demonstration purposes.", ENDPOINT))
}
```

```{r}
#| label: summary
#| output: asis
if (endpoint_available) {
  flagged <- subjects |> slice_max(dropout_risk, n = 50)
  hit <- mean(flagged$discontinued) * 100
  base <- discont_rate * 100
  cat(sprintf(
    "Sort every subject by predicted risk. The 50 highest show a **%.0f%%** discontinuation rate, against **%.1f%%** for the whole study. That is **%.1fx** the study rate, and a list of 50 is short enough for a study manager to work through site by site.\n\n",
    hit, base, hit / base
  ))
  cat("The two strongest drivers are the adverse event count and the missed visits in the first 12 weeks. Every site already records both, so no new data collection is necessary.")
}
```

## Method

A script makes the data. The data goes to Amazon S3 as parquet, and the
tables are registered in the AWS Glue Data Catalog.

This report reads those tables through Amazon Athena when it renders. It
then scores every subject against a logistic dropout-risk model. Amazon
SageMaker hosts that model on a real-time endpoint. The model itself trains
in Python (`ml/train_and_deploy.py`) and is not refit here.

Every step runs from Positron on Amazon SageMaker, and the report is
published to Posit Connect.

::: {.callout-tip collapse="true"}
## Build this report again

```bash
uv run python data/generate.py     --domain lifesci
uv run python load/load_athena.py  --domain lifesci
uv run python ml/train_and_deploy.py --domain lifesci
quarto render reports/helix_trials/enrollment_safety_review_r.qmd
uv run python ml/teardown.py --all
```
:::
