---
title: "Helix Therapeutics: study HLX-301 walkthrough (R)"
subtitle: "Enrollment and retention, one cell at a time, in tidyverse and tidymodels"
format:
  html:
    theme: cosmo
    brand: ../_brand.yml
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 9
    fig-height: 4
execute:
  warning: false
  message: false
engine: knitr
---

::: {.callout-important}
## Synthetic data

This project contains synthetic data and analysis created for demonstration
purposes only. Helix Therapeutics and study HLX-301 are fictional.
:::

This is the R parallel of `analysis/walkthrough_lifesci.qmd`. Every query and
every conclusion is the same; only the language changes. Reads go through the
`odbc` package instead of `pyodbc`, exploration uses `dplyr` and `ggplot2`
instead of `pandas` and `matplotlib`, and `tidymodels`' `yardstick` package
computes the ranking metric.

Run this file one cell at a time in Positron on Amazon SageMaker. Use
Ctrl+Enter, or Cmd+Enter on a Mac. You can also render the whole file with
`uv run quarto render analysis/walkthrough_lifesci_r.qmd`.

Study HLX-301 is a randomized, placebo-controlled Phase III trial. The
efficacy measure is a disease activity score, or DAS, where a lower score is
better.

A study team asks one question. **At Week 12, which subjects are likely to
leave the study, so that sites can act?**

## 1. No keys and no setup

```{r}
library(DBI)
library(dplyr)
library(ggplot2)
library(patchwork)
library(paws)
library(tidymodels)

find_repo_root <- function(start = getwd()) {
  path <- normalizePath(start)
  while (!file.exists(file.path(path, "config.R"))) {
    parent <- dirname(path)
    if (parent == path) stop("Could not find config.R above ", start)
    path <- parent
  }
  path
}
repo <- find_repo_root()
source(file.path(repo, "config.R"))
source(file.path(repo, "athena.R"))

sts <- paws::sts(config = list(region = athena_region()))
identity <- sts$get_caller_identity()
cat("account", identity$Account, "\n")
cat("region ", athena_region(), "\n")
```

## 2. The governed catalog

```{r}
athena_tables("helix_trials")
```

## 3. A query that looks correct

Start with the simplest question in the study. What share of subjects left
early?

A natural way to write it is to join the subject table to a visit, so that
the query reads over subjects who have a record at that visit.

```{r}
naive <- athena_query(
  "
  SELECT count(*)                                                   AS subjects,
         avg(CASE WHEN s.discontinued THEN 1.0 ELSE 0.0 END) * 100   AS discontinued_pct
  FROM dim_subject s
  JOIN fct_visit_observations o ON o.subject_id = s.subject_id
  WHERE o.visit_id = 7          -- Week 24
  ",
  "helix_trials"
)
naive
```

That number is too low, and nothing about the query looks wrong.

The join reads only over subjects with a Week 24 record. A subject who
withdrew at Week 8 has no such record, so the query never counts them. The
rows that the query needs most are the rows it removes.

```{r}
total <- athena_query(
  "
  SELECT count(*)                                                AS subjects,
         avg(CASE WHEN discontinued THEN 1.0 ELSE 0.0 END) * 100  AS discontinued_pct
  FROM dim_subject
  ",
  "helix_trials"
)
lost <- as.integer(total$subjects - naive$subjects)
cat(sprintf("joined at Week 24 : %.1f%%  over %s subjects\n",
            naive$discontinued_pct, format(naive$subjects, big.mark = ",")))
cat(sprintf("every subject     : %.1f%%  over %s subjects\n",
            total$discontinued_pct, format(total$subjects, big.mark = ",")))
cat(sprintf("\n%d subjects never reached Week 24. The first query dropped them.\n", lost))
```

This is survivorship bias. It passes review, it runs without an error, and
it reports less than half the true rate.

## 4. The query that is correct

Group the visit history over a fixed early window first, Screening to Week
12, and then join. This keeps every randomized subject, and it avoids a
second error.

A raw count over the whole study depends on exposure. A subject who leaves
early has fewer visits available to miss, so the coefficient gets the wrong
sign.

```{r}
subjects <- athena_query(
  "
  WITH early_window AS (
      SELECT subject_id,
             count(*)                                                  AS early_visits,
             sum(CASE WHEN adverse_event_flag THEN 1 ELSE 0 END)        AS early_ae,
             CAST(sum(CASE WHEN NOT visit_completed THEN 1 ELSE 0 END) AS double)
                 / count(*)                                            AS missed_visit_rate
      FROM fct_visit_observations
      WHERE visit_id <= 5
      GROUP BY subject_id
  )
  SELECT s.subject_id, s.arm, s.age, s.sex, s.ecog_status, s.bmi,
         s.prior_therapy_lines, s.baseline_das, s.early_ae_count,
         w.missed_visit_rate, s.discontinued,
         t.site_name, t.country, t.region
  FROM dim_subject s
  JOIN early_window w ON w.subject_id = s.subject_id
  JOIN dim_site t     ON t.site_id    = s.site_id
  ",
  "helix_trials"
) |>
  as_tibble()

cat(format(nrow(subjects), big.mark = ","), "subjects and", ncol(subjects), "columns\n")
cat(sprintf("discontinuation rate  %.1f%%\n", mean(subjects$discontinued) * 100))
head(subjects)
```

## 5. Open this in the Data Explorer

Click `subjects` in the Variables pane. Sort by `missed_visit_rate`, filter
the arm to Treatment, and read the column summaries. No code is necessary.

```{r}
subjects |>
  summarise(
    age = mean(age), baseline_das = mean(baseline_das),
    early_ae_count = mean(early_ae_count),
    missed_visit_rate = mean(missed_visit_rate),
    .by = arm
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

## 6. Efficacy and safety

```{r}
das <- athena_query(
  "
  SELECT v.visit_name, v.scheduled_day, s.arm,
         avg(o.das_score)      AS mean_das,
         count(*)              AS observations
  FROM fct_visit_observations o
  JOIN dim_subject s ON s.subject_id = o.subject_id
  JOIN dim_visit   v ON v.visit_id   = o.visit_id
  WHERE v.scheduled_day >= 0
  GROUP BY v.visit_name, v.scheduled_day, s.arm
  ORDER BY v.scheduled_day
  ",
  "helix_trials"
)

p1 <- ggplot(das, aes(scheduled_day, mean_das, color = arm)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  scale_color_manual(values = c(Treatment = "#447099", Placebo = "#ee6331")) +
  labs(title = "Disease activity score by arm, lower is better",
       x = "study day", y = "mean DAS") +
  theme_minimal()

by_ae <- subjects |> summarise(rate = mean(discontinued) * 100, .by = early_ae_count)

p2 <- ggplot(by_ae, aes(factor(early_ae_count), rate)) +
  geom_col(fill = "#9a4665", width = 0.72) +
  labs(title = "Discontinuation by adverse events to Week 12",
       x = "adverse events recorded by Week 12", y = "discontinued (%)") +
  theme_minimal()

p1 + p2
```

The treatment arm separates from placebo, and it holds to the end of the
study. Dropout rises quickly as early adverse events increase.

## 7. Who is still in the study?

Retention by arm, over the visit schedule.

```{r}
retention <- athena_query(
  "
  SELECT v.scheduled_day, s.arm, count(*) AS attended
  FROM fct_visit_observations o
  JOIN dim_subject s ON s.subject_id = o.subject_id
  JOIN dim_visit   v ON v.visit_id   = o.visit_id
  WHERE v.scheduled_day >= 0
  GROUP BY v.scheduled_day, s.arm
  ORDER BY v.scheduled_day
  ",
  "helix_trials"
)

retention <- retention |>
  arrange(scheduled_day) |>
  mutate(share = attended / first(attended) * 100, .by = arm)

ggplot(retention, aes(scheduled_day, share, color = arm)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  scale_color_manual(values = c(Treatment = "#447099", Placebo = "#ee6331")) +
  coord_cartesian(ylim = c(70, 101)) +
  labs(title = "Share of subjects still attending, by study day",
       x = "study day", y = "still in the study (%)") +
  theme_minimal()
```

## 8. Score the cohort with the hosted model

SageMaker hosts the dropout-risk model on a real-time endpoint.

```{r}
features <- c("age", "ecog_status", "bmi", "prior_therapy_lines",
              "baseline_das", "early_ae_count", "missed_visit_rate", "arm")

runtime <- paws::sagemakerruntime(config = list(region = athena_region()))
scored <- TRUE
domain <- get_domain("lifesci")

body <- jsonlite::toJSON(list(
  instances = subjects |> select(all_of(features))
), auto_unbox = FALSE, dataframe = "rows")

result <- tryCatch(
  runtime$invoke_endpoint(
    EndpointName = domain$endpoint,
    ContentType = "application/json",
    Body = charToRaw(body)
  ),
  error = function(err) err
)

if (inherits(result, "error")) {
  scored <- FALSE
  cat("The endpoint", domain$endpoint, "did not answer.\n")
  cat(" ", conditionMessage(result), "\n")
  cat("  Deploy it:  uv run python ml/train_and_deploy.py --domain lifesci\n")
} else {
  predictions <- jsonlite::fromJSON(rawToChar(result$Body))$predictions
  subjects <- subjects |> mutate(dropout_risk = predictions)
  cat("scored", format(nrow(subjects), big.mark = ","), "subjects on", domain$endpoint, "\n")
}
```

```{r}
if (scored) {
  p1 <- ggplot(subjects, aes(dropout_risk, fill = factor(discontinued,
                                                          labels = c("stayed in the study", "left the study")))) +
    geom_histogram(bins = 26, alpha = 0.65, position = "identity") +
    scale_fill_manual(values = c("stayed in the study" = "#447099",
                                  "left the study" = "#ee6331"), name = NULL) +
    labs(title = "Predicted risk, split by what happened",
         x = "predicted dropout risk", y = "subjects") +
    theme_minimal()

  sites <- subjects |>
    summarise(enrolled = n(), risk = mean(dropout_risk) * 100,
              actual = mean(discontinued) * 100, .by = site_name) |>
    filter(enrolled >= 15)

  p2 <- ggplot(sites, aes(risk, actual, size = enrolled)) +
    geom_point(alpha = 0.6, color = "#419599") +
    labs(title = "Site risk against site outcome",
         x = "mean predicted risk (%)", y = "discontinued so far (%)") +
    theme_minimal()

  p1 + p2
}
```

## 9. Which sites need a retention conversation?

This table is what a study manager acts on.

```{r}
if (scored) {
  watchlist <- subjects |>
    summarise(
      enrolled = n(),
      predicted_risk = mean(dropout_risk) * 100,
      discontinued_pct = mean(discontinued) * 100,
      .by = c(region, site_name)
    ) |>
    filter(enrolled >= 20) |>
    arrange(desc(predicted_risk))
  head(watchlist, 10) |> mutate(across(where(is.numeric), \(x) round(x, 1)))
}
```

```{r}
#| output: asis
if (scored) {
  flagged <- subjects |> slice_max(dropout_risk, n = 25)
  hit <- mean(flagged$discontinued) * 100
  base <- mean(subjects$discontinued) * 100
  cat(sprintf(
    "The 25 subjects with the highest predicted risk show a **%.0f%%** discontinuation rate. The rate for the whole study is **%.1f%%**.\n",
    hit, base
  ))
}
```

## 10. The experiment history

Every model that trained for this demo is a run on the SageMaker managed
MLflow tracking server. Pulling the run history into a data frame uses the
Python `mlflow` client together with `sagemaker-mlflow`'s SigV4 signer, which
has no R port today. This step confirms the server is reachable and opens the
same signed UI that `ml/tracking.py` links to.

```{r}
sm <- paws::sagemaker(config = list(region = athena_region()))

status <- tryCatch(
  sm$describe_mlflow_tracking_server(TrackingServerName = mlflow_server_name())$TrackingServerStatus,
  error = function(e) NULL
)

if (is.null(status)) {
  cat("no MLflow server named", mlflow_server_name(), "\n")
  cat("Create it:  uv run python ml/mlflow_server.py create\n")
} else if (status != "Created") {
  cat("the MLflow server is", status, "not running.\n")
  cat("Start it:  uv run python ml/mlflow_server.py start\n")
} else {
  url <- tryCatch(
    sm$create_presigned_mlflow_tracking_server_url(
      TrackingServerName = mlflow_server_name()
    )$AuthorizedUrl,
    error = function(e) NULL
  )
  cat("tracking server is running.\n")
  cat(if (!is.null(url)) url else "No signed URL. Try again in a moment.", "\n")
  cat("For the run-by-run comparison of feature sets, open",
      "analysis/walkthrough_lifesci.qmd, section 10.\n")
}
```

The comparison in the Python walkthrough answers a clinical question. What a
site knows at enrollment predicts dropout poorly. The adverse events and the
missed visits from the first 12 weeks predict it better, and the two
together are better still. Every site already records those signals.

## What comes next

```bash
uv run quarto render reports/helix_trials/enrollment_safety_review_r.qmd
uv run python ml/teardown.py --all
```
