---
title: "Aurora Lending Group: portfolio risk walkthrough (R)"
subtitle: "The live demo, one cell at a time, in tidyverse and tidymodels"
format:
  html:
    theme: cosmo
    brand: ../_brand.yml
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 9
    fig-height: 4
execute:
  warning: false
  message: false
engine: knitr
---

::: {.callout-important}
## Synthetic data

This project contains synthetic data and analysis created for demonstration
purposes only. Aurora Lending Group is a fictional company.
:::

This is the R parallel of `analysis/walkthrough_finance.qmd`. Every step reads
the same tables and calls the same hosted endpoint; only the language changes.
Reads go through the `odbc` package instead of `pyodbc`, exploration uses
`dplyr` and `ggplot2` instead of `pandas` and `matplotlib`, and the decile
metrics come from `tidymodels`' `yardstick` package.

Run this file one cell at a time in Positron on Amazon SageMaker. Use
Ctrl+Enter, or Cmd+Enter on a Mac. You can also render the whole file with
`uv run quarto render analysis/walkthrough_finance_r.qmd`.

## 1. No keys and no setup

The Positron image holds an AWS profile that asks the Studio app for
credentials. `paws` therefore finds the SageMaker execution role by itself.

```{r}
library(DBI)
library(dplyr)
library(ggplot2)
library(patchwork)
library(paws)
library(tidymodels)

# Find the repository root, so this file works from any directory.
find_repo_root <- function(start = getwd()) {
  path <- normalizePath(start)
  while (!file.exists(file.path(path, "config.R"))) {
    parent <- dirname(path)
    if (parent == path) stop("Could not find config.R above ", start)
    path <- parent
  }
  path
}
repo <- find_repo_root()
source(file.path(repo, "config.R"))
source(file.path(repo, "athena.R"))

sts <- paws::sts(config = list(region = athena_region()))
identity <- sts$get_caller_identity()
cat("account", identity$Account, "\n")
cat("role   ", sub(".*/([^/]+)/.*", "\\1", identity$Arn), "\n")
cat("region ", athena_region(), "\n")
```

## 2. The governed catalog

AWS Glue is the catalog. Athena is the engine. Nothing was copied here.

```{r}
athena_tables("aurora_lending")
```

## 3. A query, straight into a tibble

The workgroup `primary` has no default output location. Every query must
therefore pass a staging location, which `athena_odbc_string()` always
includes, or Athena returns an error.

```{r}
loans <- athena_query(
  "
  SELECT f.loan_id,
         f.origination_date,
         f.fico_at_origination,
         f.dti,
         f.apr,
         f.term_months,
         CAST(f.principal_amount AS double)    AS principal_amount,
         CAST(f.outstanding_balance AS double) AS outstanding_balance,
         f.employment_length_years,
         f.purpose,
         f.charged_off,
         b.state,
         b.income_band,
         b.homeownership,
         p.product_name
  FROM fct_loan_performance f
  JOIN dim_borrower b     ON b.borrower_id = f.borrower_id
  JOIN dim_loan_product p ON p.product_id  = f.product_id
  ",
  "aurora_lending"
) |>
  as_tibble()

cat(format(nrow(loans), big.mark = ","), "loans and", ncol(loans), "columns\n")
head(loans)
```

## 4. Open this in the Data Explorer

Click `loans` in the Variables pane. The Data Explorer then sorts any column,
filters rows, and shows a summary of every column, with no code. Take your
time on this step.

```{r}
summary(loans)
```

## 5. Where is the risk?

```{r}
loans <- loans |>
  mutate(
    fico_band = cut(
      fico_at_origination,
      breaks = c(579, 620, 660, 700, 740, 780, 850),
      labels = c("580-619", "620-659", "660-699", "700-739", "740-779", "780+")
    ),
    vintage = as.integer(format(as.Date(origination_date), "%Y"))
  )

by_band <- loans |>
  summarise(rate = mean(charged_off) * 100, .by = fico_band)

p1 <- ggplot(by_band, aes(fico_band, rate)) +
  geom_col(fill = "#447099", width = 0.72) +
  labs(title = "Charge-off rate by FICO band", x = "FICO at origination",
       y = "charge-off rate (%)") +
  theme_minimal()

by_vintage <- loans |>
  summarise(rate = mean(charged_off) * 100, .by = c(vintage, fico_band))

p2 <- ggplot(by_vintage, aes(vintage, rate, color = fico_band)) +
  geom_line(linewidth = 0.9) +
  geom_point() +
  scale_color_viridis_d(name = "FICO band") +
  labs(title = "Charge-off rate by year", x = "year of origination", y = "") +
  theme_minimal() +
  theme(legend.text = element_text(size = 7))

p1 + p2
```

Credit quality is the strongest driver. The newest years have less time on
book, so their rates are still rising.

## 6. Where is the money?

A rate alone does not size the problem. This chart puts the balance at risk
next to the rate.

```{r}
by_purpose <- loans |>
  summarise(
    rate = mean(charged_off) * 100,
    exposure = sum(outstanding_balance) / 1e6,
    .by = purpose
  ) |>
  arrange(rate) |>
  mutate(purpose_label = gsub("_", " ", purpose))

ggplot(by_purpose, aes(rate, reorder(purpose_label, rate))) +
  geom_col(fill = "#447099") +
  geom_text(aes(label = sprintf("$%sM outstanding", format(round(exposure), big.mark = ","))),
            hjust = -0.05, size = 3, color = "#404041") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.35))) +
  labs(title = "Charge-off rate and balance outstanding, by loan purpose",
       x = "charge-off rate (%)", y = NULL) +
  theme_minimal()
```

## 7. Score the book with the hosted model

SageMaker hosts the model on a real-time endpoint. This cell sends an HTTPS
request, which the same execution role authorizes. Nothing loads locally.

```{r}
features <- c("fico_at_origination", "dti", "apr", "term_months",
              "principal_amount", "employment_length_years", "purpose")

set.seed(20260821)
sample_loans <- loans |> slice_sample(n = 1000)
runtime <- paws::sagemakerruntime(config = list(region = athena_region()))
scored <- TRUE

domain <- get_domain("finance")
body <- jsonlite::toJSON(list(
  instances = sample_loans |> select(all_of(features))
), auto_unbox = FALSE, dataframe = "rows")

result <- tryCatch(
  runtime$invoke_endpoint(
    EndpointName = domain$endpoint,
    ContentType = "application/json",
    Body = charToRaw(body)
  ),
  error = function(err) err
)

if (inherits(result, "error")) {
  scored <- FALSE
  cat("The endpoint", domain$endpoint, "did not answer.\n")
  cat(" ", conditionMessage(result), "\n")
  cat("  Deploy it:  uv run python ml/train_and_deploy.py --domain finance\n")
} else {
  predictions <- jsonlite::fromJSON(rawToChar(result$Body))$predictions
  sample_loans <- sample_loans |> mutate(risk_score = predictions)
  cat("scored", format(nrow(sample_loans), big.mark = ","), "loans on", domain$endpoint, "\n")
}
```

```{r}
if (scored) {
  sample_loans |>
    select(loan_id, fico_at_origination, dti, purpose, risk_score, charged_off) |>
    head(10)
}
```

## 8. Does the model rank the risk?

Sort by predicted risk, cut into ten equal groups, then compare the
prediction against the charge-off rate that each group reached.
`tidymodels`' `yardstick` package computes the overall AUC; the decile table
adds a business-readable view of the same separation.

```{r}
if (scored) {
  sample_loans <- sample_loans |>
    mutate(
      decile = ntile(risk_score, 10),
      charged_off_fct = factor(charged_off, levels = c(1, 0))
    )

  auc <- sample_loans |>
    yardstick::roc_auc(truth = charged_off_fct, risk_score, event_level = "first")
  cat("test AUC (yardstick):", round(auc$.estimate, 3), "\n")

  lift <- sample_loans |>
    summarise(
      loans = n(),
      mean_predicted = mean(risk_score) * 100,
      actual_chargeoff = mean(charged_off) * 100,
      exposure = sum(outstanding_balance),
      .by = decile
    ) |>
    arrange(decile)
  lift
}
```

```{r}
if (scored) {
  lift_long <- lift |>
    tidyr::pivot_longer(c(mean_predicted, actual_chargeoff),
                         names_to = "series", values_to = "rate")

  p1 <- ggplot(lift_long, aes(factor(decile), rate, fill = series)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    scale_fill_manual(values = c(mean_predicted = "#447099", actual_chargeoff = "#ee6331"),
                       labels = c("predicted", "actual")) +
    labs(title = "Predicted rate against actual rate, by decile",
         x = "risk decile, 10 is the riskiest", y = "charge-off rate (%)", fill = NULL) +
    theme_minimal()

  # A model that ranks well puts most of the losses in the top deciles.
  ordered <- lift |> arrange(desc(decile)) |>
    mutate(
      captured = cumsum(actual_chargeoff / 100 * loans),
      captured = captured / max(captured) * 100,
      share = cumsum(loans) / sum(loans) * 100
    )

  p2 <- ggplot(ordered, aes(share, captured)) +
    geom_line(color = "#9a4665", linewidth = 1) +
    geom_point(color = "#9a4665") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "#a0a0a0") +
    labs(title = "Share of all charge-offs captured",
         x = "share of the book reviewed (%)", y = "share of charge-offs found (%)") +
    theme_minimal()

  p1 + p2
}
```

```{r}
#| output: asis
if (scored) {
  base <- mean(sample_loans$charged_off) * 100
  top <- lift$actual_chargeoff[lift$decile == 10]
  two <- sum(sample_loans$charged_off[sample_loans$decile >= 9]) /
    max(sum(sample_loans$charged_off), 1) * 100
  cat(sprintf(
    "The base rate for this sample is **%.1f%%**. The riskiest decile reaches **%.1f%%**, which is **%.1fx** the base rate. The top two deciles hold **%.0f%%** of all charge-offs.\n",
    base, top, top / base, two
  ))
}
```

## 9. The experiment history

Every model that trained for this demo is a run on the SageMaker managed
MLflow tracking server. Pulling the run history into a data frame uses the
Python `mlflow` client together with `sagemaker-mlflow`, which signs each
request with SigV4; there is no R port of that signer today. This step
instead confirms the server is reachable and opens the same signed UI that
`ml/tracking.py` links to.

```{r}
sm <- paws::sagemaker(config = list(region = athena_region()))

status <- tryCatch(
  sm$describe_mlflow_tracking_server(TrackingServerName = mlflow_server_name())$TrackingServerStatus,
  error = function(e) NULL
)

if (is.null(status)) {
  cat("no MLflow server named", mlflow_server_name(), "\n")
  cat("Create it:  uv run python ml/mlflow_server.py create\n")
} else if (status != "Created") {
  cat("the MLflow server is", status, "not running.\n")
  cat("Start it:  uv run python ml/mlflow_server.py start\n")
} else {
  url <- tryCatch(
    sm$create_presigned_mlflow_tracking_server_url(
      TrackingServerName = mlflow_server_name()
    )$AuthorizedUrl,
    error = function(e) NULL
  )
  cat("tracking server is running.\n")
  cat(if (!is.null(url)) url else "No signed URL. Try again in a moment.", "\n")
  cat("For the run-by-run comparison of feature sets, open",
      "analysis/walkthrough_finance.qmd, section 9.\n")
}
```

## What comes next

The same query and the same endpoint drive the published report.

```bash
uv run quarto render reports/aurora_lending/portfolio_risk_review_r.qmd
```

After the session, stop the endpoint billing.

```bash
uv run python ml/teardown.py --all
```
